---
title: "rcicrdiagnostics: a tutorial"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{rcicrdiagnostics: a tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.path = "tutorial-"
)
set.seed(1)
```

## 1. What this package is

Reverse correlation (RC) experiments let you visualise a participant's
mental representation — for example, what their mental image of
"trustworthy" looks like — by averaging the noise patterns the
participant chose across many trials. The end product is a
**classification image (CI)**, and a common summary of the CI's quality
is the **information value (infoVal)**: roughly, how much real signal is
in the CI versus chance (Brinkman et al., 2017, 2019).

These pipelines can fail quietly. A response column coded `{0, 1}`
instead of `{-1, 1}`, a stimulus-number column that is off by one, a
noise-matrix file whose dimensions don't match the response data — any
of these produces a CI that *looks* plausible but is wrong. R doesn't
catch these problems and neither does the `rcicr` package, because the
errors live in the *match* between your response data and the files that
generated the stimuli.

`rcicrdiagnostics` is a set of checks you run on your data **before**
computing any CI or infoVal. If a check fails, fix the data; if it
warns, investigate; if everything passes, you can compute CIs knowing
the basic data mechanics are in order.

The package supports two pipelines through a single `method` argument:

- **`"2ifc"`** — the standard two-image forced-choice pipeline, as
  implemented by the [`rcicr`][rcicr] package (Dotsch, 2016). On each
  trial the participant picks one of two noisy faces.
- **`"briefrc"`** — the Brief Reverse Correlation pipeline
  (Schmitz et al., 2024). On each trial the participant picks one noisy
  face out of 6 or 12. CIs are computed by direct noise-matrix
  multiplication rather than by reconstructing sinusoids from a
  generating `.RData` file (so there is no `.RData` file for briefRC).

You only need to know *your* pipeline to use this package. The checks
use the right logic automatically once you set `method`.

[rcicr]: https://github.com/rdotsch/rcicr

## 2. Installation and requirements

### R version

R 4.1 or newer.

### Packages installed automatically

`rcicrdiagnostics` has only two hard dependencies, both on CRAN:

- [`data.table`](https://rdatatable.gitlab.io/data.table/) — a fast
  alternative to base R's `data.frame`, used internally for counting
  and grouping. You don't need to learn its syntax to use this package.
- [`cli`](https://cli.r-lib.org/) — used to format console messages
  with colour.

### Optional packages

- [`rcicr`](https://github.com/rdotsch/rcicr) — only needed once the
  CI-generation–dependent checks land (see Section 9). Install from
  GitHub:

    ```r
    remotes::install_github("rdotsch/rcicr")
    ```

- `testthat`, `knitr`, `rmarkdown`, `withr` — only needed to build the
  package or its vignettes locally.

### Installing the package

`rcicrdiagnostics` is distributed from GitHub (not CRAN):

```r
# install.packages("remotes")
remotes::install_github("olivethree/rcicrdiagnostics")
```

Or with `pak`:

```r
# install.packages("pak")
pak::pak("olivethree/rcicrdiagnostics")
```

Confirm it loaded:

```{r load, message = FALSE}
library(rcicrdiagnostics)
packageVersion("rcicrdiagnostics")
```

## 3. What data the package expects

### Response data (both pipelines)

A trial-level data frame — one row per trial — that you either load
from a CSV or build in R. At minimum it needs:

| Column           | Purpose                                                             |
| ---------------- | ------------------------------------------------------------------- |
| `participant_id` | Identifier for each participant. Character or integer is fine.      |
| `stimulus`       | The stimulus number for that trial. Must match the numbering used when the stimuli were generated. |
| `response`       | The participant's response on that trial. Coding depends on method. |
| `rt` (optional)  | Response time in milliseconds. Used by future RT checks.            |

If your column names differ, every function accepts overrides via
`col_participant`, `col_stimulus`, `col_response`, and `col_rt`.

### Response coding by method

- **`"2ifc"`**: `response` must be numeric `{-1, 1}`. The sign matters
  because the CI is a weighted sum of the per-trial noise patterns:
  a `+1` means the noise on that trial is added to the CI, a `-1` means
  it's subtracted. If responses are coded `{0, 1}` instead, the "subtract"
  information is lost and the CI comes out close to blank. This is one
  of the most common silent failures in RC data; Section 6.1 detects it.
- **`"briefrc"`**: `response` may be `{-1, 1}` **or** continuous weights
  (e.g. one-hot per trial, or aggregated per-stimulus weights). Any
  finite numeric value is accepted.

### Auxiliary inputs by method

- **2IFC** uses the `.RData` file produced by
  `rcicr::generateStimuli2IFC()` plus the base image (the neutral face
  the noise was added to). Pass these as `rdata` and `base_image`.
- **Brief-RC** uses a **noise matrix**: a plain-text file with pixels
  as rows and stimuli as columns, each cell a real-valued noise weight.
  For 128×128 images and 300 stimuli, the matrix is 16 384 rows ×
  300 columns. Pass as `noise_matrix`.

## 4. A first diagnostic run

Here is a small synthetic 2IFC dataset — three participants, one hundred
trials each, responses drawn uniformly from `{-1, 1}`:

```{r toy-2ifc}
responses <- data.frame(
  participant_id = rep(1:3, each = 100),
  stimulus       = rep(1:100, 3),
  response       = sample(c(-1, 1), 300, replace = TRUE)
)
head(responses)
```

Run the full set of checks:

```{r run-2ifc}
report <- run_all_checks(responses, method = "2ifc")
report
```

That printout is what the package produces for you: one line per check,
a coloured tag (green `[PASS]`, yellow `[WARN]`, red `[FAIL]`, grey
`[SKIP]` on colour-capable consoles), and a short explanation.

## 5. The result object

Every single check returns a small list with a fixed shape. The package
gives this shape a class name — `rcdiag_result` — so that printing and
other helpers know how to handle it:

```{r inspect-result}
r <- check_response_coding(responses, method = "2ifc")
str(r, max.level = 1)
```

Four fields you can read directly:

- `status` — one of `"pass"`, `"warn"`, `"fail"`, `"skip"`. Interpret as:
  move on, investigate, stop and fix, not run.
- `label` — short description of the check.
- `detail` — the lines printed under the status tag.
- `data` — optional supporting data: flagged participants, count tables,
  group statistics. Useful when you want to programmatically exclude
  problem participants from the dataset you eventually feed to the CI
  generator.

`run_all_checks()` returns a collection of these, wrapped in an
`rcdiag_report`. `summary()` flattens the collection into a tidy data
frame — handy if you want to filter or log statuses:

```{r summary-report}
summary(report)
```

## 6. The checks one by one

Every example below uses synthetic data built inline, so you can run
the whole tutorial start to finish without any external files.

### 6.1. `check_response_coding()`

Verifies the response column is coded as `method` expects. For 2IFC it
detects the two most common miscodings and suggests the one-line fix.

```{r coding-ok}
ok <- data.frame(
  participant_id = 1:10,
  stimulus       = 1:10,
  response       = sample(c(-1, 1), 10, replace = TRUE)
)
check_response_coding(ok, method = "2ifc")
```

```{r coding-miscode}
miscoded <- data.frame(
  participant_id = 1:10,
  stimulus       = 1:10,
  response       = sample(c(0, 1), 10, replace = TRUE)
)
check_response_coding(miscoded, method = "2ifc")
```

The `{0, 1}` miscoding returns **warn** with the recoding formula
(`responses$response <- responses$response * 2 - 1`). The `{1, 2}`
miscoding returns the analogous fix. For `method = "briefrc"`, any
finite numeric passes; `NA` responses produce a warning so you can
decide whether to drop or impute them.

### 6.2. `check_trial_counts()`

Counts trials per participant. If you pass `expected_n`, it flags
participants whose count doesn't match. `expected_n` can be a single
number (everyone should have the same number of trials) or a vector of
allowed counts (for designs with multiple trial-count conditions, e.g.
`c(300, 500, 1000)`).

```{r counts-pass}
three_hundred_each <- data.frame(
  participant_id = rep(1:3, each = 300),
  stimulus       = rep(1:300, 3),
  response       = 1L
)
check_trial_counts(three_hundred_each, expected_n = 300)
```

```{r counts-warn}
with_a_short_participant <- data.frame(
  participant_id = rep(1:5, times = c(300, 300, 300, 300, 250)),
  stimulus       = 1,
  response       = 1L
)
check_trial_counts(with_a_short_participant, expected_n = 300)
```

A single off-count participant is **warn**; more than 10% of
participants off-count is **fail** (the logic: one odd participant may
be excludable, but a systemic mismatch means your merge or filter logic
is probably wrong). Leave `expected_n = NULL` (the default) to get a
count distribution without flagging anything.

### 6.3. `check_duplicates()`

Duplicate rows almost always indicate a merge bug (for example, two
partial exports of the same participant concatenated together). Two
cases are flagged separately:

1. **Fully duplicated rows** — every column identical. Unambiguously bad.
2. **`(participant, stimulus)` pair duplicates** — same participant,
   same stimulus, but different response or RT. Could be legitimate if
   your design deliberately repeats stimuli, or could be a merge error.

```{r dup}
with_a_dup <- data.frame(
  participant_id = c(1, 1, 1, 2, 2),
  stimulus       = c(1, 2, 2, 1, 2),
  response       = c(1, -1, -1, 1, -1)
)
check_duplicates(with_a_dup)
```

### 6.4. `check_response_bias()`

Two kinds of bias are reported:

- **Constant responders** — a participant who gave the same response on
  every trial. These are almost always bad data (disengaged participant
  or a coding bug) because their "signal" contribution is constant and
  does nothing except scale the base image. Status: **fail**.
- **Extremely biased participants** — those whose mean response is far
  enough from zero that you probably can't trust the CI. The threshold
  is `|mean| > bias_threshold`, default `0.6`. For binary `{-1, 1}`
  coding this corresponds to roughly an 80/20 split. Status: **warn**.

```{r bias}
with_a_constant_responder <- data.frame(
  participant_id = rep(1:4, each = 100),
  stimulus       = rep(1:100, 4),
  response       = c(
    sample(c(-1, 1), 100, replace = TRUE),
    sample(c(-1, 1), 100, replace = TRUE),
    rep(1, 100),
    sample(c(-1, 1), 100, replace = TRUE)
  )
)
check_response_bias(with_a_constant_responder)
```

The per-participant breakdown is stored on the result so you can filter
your dataset directly without re-computing:

```{r bias-data}
r <- check_response_bias(with_a_constant_responder)
r$data$per_participant
```

### 6.5. Noise matrix (Brief-RC)

`read_noise_matrix()` reads a space-delimited text file into a numeric
matrix. For a self-contained example we just build one in memory and
validate it:

```{r noise}
mat <- matrix(rnorm(16384 * 5, sd = 0.05), nrow = 16384, ncol = 5)
validate_noise_matrix(mat, expected_pixels = 16384, expected_stimuli = 5)
```

The validator checks the dimensions (one row per pixel, one column per
stimulus) and that no cell is `NA`, `NaN`, or `Inf`. Supplying the wrong
expected dimensions returns **fail** with a specific message. A lot of
early-pipeline errors surface here — confusing pixel count with stimulus
count, saving the matrix with an unexpected delimiter, and so on.

### 6.6. Loading response data from disk

`load_responses()` is a thin wrapper around `data.table::fread()` (a
fast CSV reader) that checks the required columns are present. Use it
on your real data file:

```r
responses <- load_responses("my_data.csv", method = "2ifc")
```

It returns a `data.table`, which behaves like a regular `data.frame`
for the check functions — you don't need to learn a new syntax.

## 7. `run_all_checks()`

`run_all_checks()` runs every implemented check in one call and returns
the collected `rcdiag_report`. It works out the `method` automatically
when exactly one of `rdata` or `noise_matrix` is supplied, so you
usually don't need to set `method` yourself:

```{r run-auto, eval = FALSE}
# Auto-detected as "2ifc" because rdata is supplied:
run_all_checks(responses, rdata = "stimuli.RData")

# Auto-detected as "briefrc":
run_all_checks(responses, noise_matrix = "noise_matrix_128.txt")
```

Pass `method` explicitly when you have neither file on hand — for
example, to sanity-check the response CSV by itself:

```{r run-explicit}
run_all_checks(responses, method = "2ifc")
```

Per-check options are passed through:

```{r run-threshold}
run_all_checks(
  responses,
  method = "2ifc",
  expected_n = 100
)
```

For programmatic access, the individual results live under
`report$results`:

```{r results-list}
report <- run_all_checks(responses, method = "2ifc")
vapply(report$results, `[[`, character(1), "status")
```

## 8. Reading the report

Four statuses:

- **pass** (green) — the check saw nothing unusual. Continue.
- **warn** (yellow) — something worth investigating but not necessarily
  fatal (e.g. one duplicate row, a known miscoding with a suggested
  fix). Look at `$data` and decide.
- **fail** (red) — something that will corrupt downstream CIs or infoVal
  if you ignore it. Fix before proceeding.
- **skip** (grey) — the check didn't run. In this release that means
  "not yet implemented" — see Section 9.

A practical workflow:

1. Run `run_all_checks()` once on the raw data.
2. Resolve every **fail** first.
3. Investigate every **warn**. Either fix it, or add a short note in
   your analysis script saying why the warning is acceptable for this
   dataset (so you or a collaborator don't re-investigate later).
4. Only once every check is **pass** — or a consciously accepted
   **warn** — move on to `rcicr::generateCI2IFC()` (2IFC) or your
   Brief-RC CI code.

## 9. Checks not yet implemented

Six checks are defined as named functions but not yet implemented.
Calling them directly stops with an informative error, and
`run_all_checks()` lists them under `report$skipped_checks` without
trying to run them.

- `check_response_inversion()` — compute the CI twice (with original
  and flipped response codes) and compare infoVal. Flags data that
  looks inverted.
- `check_stimulus_alignment()` — verify stimulus numbers line up with
  the parameters in the `.RData` (2IFC) or the columns of the noise
  matrix (briefRC).
- `check_rt()` — RT distributional checks (implausibly fast responses,
  very low within-participant variability).
- `check_version_compat()` — the `rcicr` version that produced the
  `.RData` versus the installed version.
- `compute_infoval_summary()` — per-participant infoVal with descriptive
  statistics and flags (Mangini & Biederman, 2004; Brinkman et al., 2019).
- `cross_validate_rt_infoval()` — correlate infoVal with RT quality
  metrics to spot fast-and-confident false positives.

These all depend on CI generation: via `rcicr` for 2IFC, or a
permutation-based infoVal for briefRC. Both are non-trivial and are
deliberately postponed to a later release rather than shipped as
potentially-wrong statistics.

## 10. References

Brinkman, L., Goffin, S., van de Schoot, R., van Haren, N. E., Dotsch, R., & Aarts, H. (2019). Quantifying the informational value of classification images. *Behavior Research Methods*, *51*(5), 2059–2073. https://doi.org/10.3758/s13428-019-01232-2

Brinkman, L., Todorov, A., & Dotsch, R. (2017). Visualising mental representations: A primer on noise-based reverse correlation in social psychology. *European Review of Social Psychology*, *28*(1), 333–361. https://doi.org/10.1080/10463283.2017.1381469

Dotsch, R. (2016). *rcicr: Reverse-correlation image-classification toolbox* [R package]. https://github.com/rdotsch/rcicr

Mangini, M. C., & Biederman, I. (2004). Making the ineffable explicit: Estimating the information employed for face classifications. *Cognitive Science*, *28*(2), 209–226. https://doi.org/10.1016/j.cogsci.2003.11.004

Schmitz, M., Rougier, M., & Yzerbyt, V. (2024). Introducing the brief reverse correlation: An improved tool to assess visual representations. *European Journal of Social Psychology*. Advance online publication. https://doi.org/10.1002/ejsp.3100

## 11. Authorship

Developed by M. Oliveira with Claude (Anthropic) assistance.
M. Oliveira is responsible for the content.
