---
title: "rcicrdiagnostics: a tutorial"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{rcicrdiagnostics: a tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.path = "tutorial-"
)
set.seed(1)
```

## 1. What this package is (and is not)

Reverse correlation (RC) experiments produce classification images (CIs)
and information values (infoVal) that summarise what mental representation
a participant (or group) holds. These pipelines fail **silently**: a
miscoded response, a misaligned stimulus number, or a mismatched `.RData`
file produces CIs that look plausible but are wrong. Statistical reviewers
and journal readers cannot see these problems.

`rcicrdiagnostics` is a pre-flight data-quality battery. Run it **before**
you compute a single CI. If a check fails, fix the data; if a check warns,
investigate; if every check passes, you can proceed to CI generation and
infoVal analysis with confidence that the mechanics are sound.

The package **is** a diagnostic tool for already-collected RC data. It is
not:

- an implementation of CI generation (use [`rcicr`][rcicr] for that),
- an experiment-generation toolkit,
- a general-purpose data-validation framework.

Two RC pipelines are supported through a single `method` argument:

- **`"2ifc"`** — standard two-image forced-choice, the pipeline implemented
  by [`rcicr`][rcicr].
- **`"briefrc"`** — the Brief-RC pipeline of Schmitz et al. (2024), which
  presents 6 or 12 noisy faces per trial and uses direct noise-matrix
  multiplication instead of sinusoid reconstruction.

[rcicr]: https://github.com/rdotsch/rcicr

## 2. Installation and requirements

### R version

R 4.1 or newer.

### Required packages (installed automatically)

`rcicrdiagnostics` imports only two packages, both on CRAN:

- [`data.table`](https://rdatatable.gitlab.io/data.table/) — fast data
  manipulation and CSV / noise-matrix reading.
- [`cli`](https://cli.r-lib.org/) — formatted console messages and errors.

### Optional packages

- [`rcicr`](https://github.com/rdotsch/rcicr) — needed for the 2IFC
  infoVal and CI-generation–dependent checks (not included in this
  release, see Section 9). Install from GitHub:

    ```r
    remotes::install_github("rdotsch/rcicr", ref = "development")
    ```

- `testthat`, `knitr`, `rmarkdown`, `withr` — only needed if you are
  building the package or its tests locally.

### Installing the package

`rcicrdiagnostics` is distributed from GitHub:

```r
# install.packages("remotes")
remotes::install_github("olivethree/rcicrdiagnostics")
```

Or with `pak`:

```r
# install.packages("pak")
pak::pak("olivethree/rcicrdiagnostics")
```

To confirm the install, load the package and print its version:

```{r load, message = FALSE}
library(rcicrdiagnostics)
packageVersion("rcicrdiagnostics")
```

## 3. The data the package expects

### Response data (both pipelines)

A trial-level data frame (or CSV you will read) with at minimum these
columns:

| Column           | Purpose                                                   |
| ---------------- | --------------------------------------------------------- |
| `participant_id` | Identifier for each participant. Character or integer.    |
| `stimulus`       | Stimulus number for that trial (indexes into noise/RData). |
| `response`       | The participant's response. Coding depends on `method`.   |
| `rt` (optional)  | Response time in milliseconds. Used by future RT checks.  |

Column names can be overridden via `col_participant`, `col_stimulus`,
`col_response`, and `col_rt` in every function.

### Response coding by method

- **`"2ifc"`**: `response` must be numeric `{-1, 1}`. The two common
  miscodes — `{0, 1}` and `{1, 2}` — are detected explicitly with a
  suggested fix. See Section 6.1.
- **`"briefrc"`**: `response` may be `{-1, 1}` **or** continuous weights
  (e.g. the one-hot choice, or aggregated per-stimulus weights). Any
  finite numeric is accepted.

### Auxiliary inputs by method

- **2IFC** needs the `.RData` file from `rcicr::generateStimuli2IFC()`
  (argument: `rdata`) and the base image file.
- **Brief-RC** needs a noise-matrix text file (argument: `noise_matrix`)
  with pixels as rows and stimuli as columns, and a base image file.
  For 128×128 images and 300 stimuli the matrix is 16 384 rows ×
  300 columns.

## 4. Your first diagnostic run

Here is a synthetic 2IFC dataset — three participants, one hundred trials
each, responses drawn uniformly from `{-1, 1}`:

```{r toy-2ifc}
responses <- data.frame(
  participant_id = rep(1:3, each = 100),
  stimulus       = rep(1:100, 3),
  response       = sample(c(-1, 1), 300, replace = TRUE)
)
head(responses)
```

Run the full battery:

```{r run-2ifc}
report <- run_all_checks(responses, method = "2ifc")
report
```

That output is the primary artefact of the package: one line per check,
colour-coded by status, plus an inventory of checks that are not yet
implemented.

## 5. The result object: `rcdiag_result`

Every individual check returns an object of class `rcdiag_result`. You
can inspect the underlying list:

```{r inspect-result}
r <- check_response_coding(responses, method = "2ifc")
str(r, max.level = 1)
```

Four fields:

- `status` — one of `"pass"`, `"warn"`, `"fail"`, `"skip"`. Interpret
  as: move on, investigate, stop and fix, not run.
- `label` — short one-liner describing the check.
- `detail` — character vector of diagnostic lines to show the user.
- `data` — optional list of supporting data (flagged participants, count
  tables, group statistics). Structure depends on the check.

`rcdiag_report` is the same idea for a collection of results: it is what
`run_all_checks()` returns. `summary()` produces a tidy data frame:

```{r summary-report}
summary(report)
```

## 6. Check-by-check reference

This section walks through each implemented check. Every example uses
synthetic data built inline so you can run the vignette end-to-end.

### 6.1. `check_response_coding()`

Verifies that `response` uses the coding expected by `method`. For
`"2ifc"` it detects two common miscodes explicitly:

```{r coding-ok}
ok <- data.frame(
  participant_id = 1:10,
  stimulus       = 1:10,
  response       = sample(c(-1, 1), 10, replace = TRUE)
)
check_response_coding(ok, method = "2ifc")
```

```{r coding-miscode}
miscoded <- data.frame(
  participant_id = 1:10,
  stimulus       = 1:10,
  response       = sample(c(0, 1), 10, replace = TRUE)
)
check_response_coding(miscoded, method = "2ifc")
```

The `{0, 1}` miscoding returns a **warn** with a concrete fix
(`responses$response <- responses$response * 2 - 1`). The `{1, 2}`
miscoding returns the analogous fix for its coding.

For `method = "briefrc"`, any finite numeric passes; `NA` values produce
a warning.

### 6.2. `check_trial_counts()`

Counts trials per participant and, if `expected_n` is supplied, flags
participants whose count does not match. `expected_n` can be a scalar
(every participant should have this many trials) or a vector (designs
with multiple trial-count conditions, e.g. `c(300, 500, 1000)`).

```{r counts-pass}
three_hundred_each <- data.frame(
  participant_id = rep(1:3, each = 300),
  stimulus       = rep(1:300, 3),
  response       = 1L
)
check_trial_counts(three_hundred_each, expected_n = 300)
```

```{r counts-warn}
with_a_short_participant <- data.frame(
  participant_id = rep(1:5, times = c(300, 300, 300, 300, 250)),
  stimulus       = 1,
  response       = 1L
)
check_trial_counts(with_a_short_participant, expected_n = 300)
```

The cutoff is graduated: a single off-count participant produces **warn**;
more than 10% off-count produces **fail**. Pass `expected_n = NULL`
(default) to get a distribution report without flagging.

### 6.3. `check_duplicates()`

Flags two distinct kinds of duplicates:

1. **Fully duplicated rows** — every column identical. Almost always a
   merge bug.
2. **`(participant, stimulus)` pair duplicates** — same participant, same
   stimulus, but different response or RT. May be legitimate (design
   repeats) or a pipeline gluing two runs for one participant.

```{r dup}
with_a_dup <- data.frame(
  participant_id = c(1, 1, 1, 2, 2),
  stimulus       = c(1, 2, 2, 1, 2),
  response       = c(1, -1, -1, 1, -1)
)
check_duplicates(with_a_dup)
```

### 6.4. `check_response_bias()`

Two kinds of bias:

- **Constant response** — a participant whose every trial has the same
  value. Status: **fail**.
- **Extreme mean response** — a participant whose `|mean(response)|`
  exceeds `bias_threshold` (default `0.6`, equivalent to roughly 80/20 for
  binary coding). Status: **warn**.

```{r bias}
with_a_constant_responder <- data.frame(
  participant_id = rep(1:4, each = 100),
  stimulus       = rep(1:100, 4),
  response       = c(
    sample(c(-1, 1), 100, replace = TRUE),
    sample(c(-1, 1), 100, replace = TRUE),
    rep(1, 100),
    sample(c(-1, 1), 100, replace = TRUE)
  )
)
check_response_bias(with_a_constant_responder)
```

The `data$per_participant` element of the returned object contains the
full per-participant breakdown — trial counts, means, and flags — which
you can use directly to exclude problematic participants:

```{r bias-data}
r <- check_response_bias(with_a_constant_responder)
r$data$per_participant
```

### 6.5. Noise-matrix utilities (Brief-RC)

Use `read_noise_matrix()` to read a space-delimited txt file into a
numeric matrix. For demonstration, we build a small matrix and validate
it:

```{r noise}
mat <- matrix(rnorm(16384 * 5, sd = 0.05), nrow = 16384, ncol = 5)
validate_noise_matrix(mat, expected_pixels = 16384, expected_stimuli = 5)
```

Supplying the wrong expected dimensions returns **fail** with a specific
message. Non-finite entries (`NA`, `NaN`, `Inf`) also fail.

### 6.6. Loading response data from disk

`load_responses()` is a thin wrapper around `data.table::fread()` that
validates column presence. Use it on your real data file:

```r
responses <- load_responses("my_data.csv", method = "2ifc")
```

It returns a `data.table`. You can pass that directly to any check
function (a `data.table` is also a `data.frame`).

## 7. The `run_all_checks()` orchestrator

`run_all_checks()` dispatches the full battery and returns an
`rcdiag_report`. It auto-detects `method` when you supply exactly one of
`rdata` or `noise_matrix`:

```{r run-auto, eval = FALSE}
# Auto-detected as "2ifc" because rdata is supplied:
run_all_checks(responses, rdata = "stimuli.RData")

# Auto-detected as "briefrc":
run_all_checks(responses, noise_matrix = "noise_matrix_128.txt")
```

Pass `method` explicitly when you do not have either file on hand (for
example, for a quick sanity check on response data alone):

```{r run-explicit}
run_all_checks(responses, method = "2ifc")
```

Forward per-check options through the orchestrator:

```{r run-threshold}
run_all_checks(
  responses,
  method = "2ifc",
  expected_n = 100
)
```

The report's `$results` element gives you programmatic access:

```{r results-list}
report <- run_all_checks(responses, method = "2ifc")
vapply(report$results, `[[`, character(1), "status")
```

## 8. Interpreting the report

Four statuses:

- **pass** (green) — the check saw nothing unusual. Continue.
- **warn** (yellow) — the check saw something worth investigating but
  not always fatal (e.g. one duplicate row, a known miscoding with a
  suggested fix). Look at `$data` and decide.
- **fail** (red) — the check saw something that will corrupt downstream
  CIs or infoVal if you ignore it. Fix before proceeding.
- **skip** (grey) — the check did not run. Currently reserved; the
  `skipped_checks` list in the report names the checks that are not yet
  implemented in this release.

A principled workflow:

1. Run `run_all_checks()` once on the raw data.
2. Resolve every **fail** before anything else.
3. Investigate every **warn** — either fix, or record an explicit note
   in your analysis script saying why the warn is acceptable.
4. Only after every check is **pass** (or consciously accepted **warn**),
   proceed to `rcicr::generateCI2IFC()` (2IFC) or your Brief-RC CI code.

## 9. Checks not yet implemented

Six checks are in the package as named functions but are not yet
implemented. Calling them directly aborts with an informative error,
and `run_all_checks()` skips them and lists them under
`report$skipped_checks`.

- `check_response_inversion()` — compares infoVal with original vs
  flipped response codes; flags data that looks inverted.
- `check_stimulus_alignment()` — verifies stimulus numbers align with
  the `.RData` parameters (2IFC) or noise-matrix columns (briefRC).
- `check_rt()` — RT distributional checks (too-fast responders,
  low within-participant variability).
- `check_version_compat()` — rcicr version of `.RData` vs installed
  rcicr version.
- `compute_infoval_summary()` — batch per-participant infoVal with
  descriptive statistics.
- `cross_validate_rt_infoval()` — correlate infoVal with RT quality
  metrics to detect fast-and-confident false positives.

These depend on CI generation (via `rcicr` for 2IFC, or a standalone
permutation-based infoVal for briefRC) which is a separate implementation
effort. Progress is tracked in the package's GitHub issues.

## 10. Next steps

- **Real example data**: the `data-raw/` directory contains stubs for
  `generate_example_2ifc.R` and `generate_example_briefrc.R`. Running
  these will populate `inst/extdata/` with a reproducible RC dataset
  that you can practise on.
- **Pipeline-specific vignettes**: once the CI-generation-dependent
  checks land, separate `quickstart-2ifc` and `quickstart-briefrc`
  vignettes will walk through the full analysis pipeline end-to-end.
- **A Quarto ebook**: this tutorial is written in R Markdown and maps
  cleanly to Quarto; a standalone ebook version is planned.

## 11. References

- Dotsch, R. (2016). `rcicr`: Reverse-correlation image-classification
  toolbox. <https://github.com/rdotsch/rcicr>
- Schmitz, M., et al. (2024). Brief reverse correlation.
- Mangini, M. C., & Biederman, I. (2004). Making the ineffable explicit:
  Estimating the information employed for face classifications.
  *Cognitive Science*, 28(2), 209–226.
